---
title: "Storm Analysis Coursera Project 2 for Reproducible Research"
author: "AragonAyalaCJ"
date: "2025-01-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysis of Severe Weather Events Impact on Public Health and Economy in the United States

This project look at the NOAA Storm Database (1950 - 2011) to identify the types of severe weather events that are most harmful to population health and have the greatest economic consequences.  The analysis includes data processing steps, results with visualizations, and a discussion of the findings.

## Data Processing

First, load packages.

```{r}
library(tidyverse)
```

Download file

```{r}

url_data <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

data_path <- file.path("data","StormData.csv.bz2")

if (!dir.exists("data")) {
     dir.create("data")   
} 

download.file(url_data, data_path)
```

Import data to R

```{r}
# Readr package automatically unzip bz2

raw_data <- read_csv(data_path)

```

Transform and clean data

```{r}
# Convert date columns

data <- raw_data |> 
        mutate(BGN_DATE = strptime(BGN_DATE, format = "%m/%d/%Y %H:%M:%S")) 
        
# Filter relevant columns
data <- data |> 
  select(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)

# Handle missing values
data <- na.omit(data)

# Creating a function to convert exponent symbols to numeric values

convert_exponent <- function(exp) {
  if (sum(exp %in% c("K", "k"))) {
    return(1e3)
  } else if (sum(exp %in% c("M", "m"))) {
    return(1e6)
  } else if (sum(exp %in% c("B", "b"))) {
    return(1e9)
  } else if (sum(exp %in% c("H", "h"))) {
    return(1e2)
  } else if (sum(exp %in% c("", "+", "-", "?", "0"))) {
    return(1)
  } else {
    return(as.numeric(exp))
  }
}

# Create relevant variables

data <- data |> 
        mutate(numPROPDMGEXP = convert_exponent(PROPDMGEXP),
               numCROPDMGEXP = convert_exponent(CROPDMGEXP),
               totalPROPDMGE = numPROPDMGEXP * PROPDMG,
               totalCROPDMGE = numCROPDMGEXP * CROPDMG,
               totalEconomicDmg = totalPROPDMGE + totalCROPDMGE)



```

## Results 

Most Harmful Events to Population Health (Fatalities)

```{r}
# Aggregate data by event type
health_impact <- data  |> 
  group_by(EVTYPE)  |> 
  summarize(Total_Fatalities = sum(FATALITIES), Total_Injuries = sum(INJURIES))  |> 
  arrange(desc(Total_Fatalities))

# Plot the results
ggplot(health_impact[1:10, ], aes(x = reorder(EVTYPE, -Total_Fatalities), y = Total_Fatalities)) +
  geom_bar(stat = "identity") +
  labs(title = "Top 10 Most Harmful Events to Population Health (Fatalities)",
       x = "Event Type", y = "Total Fatalities") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Most Harmful Events to Population Health (Injuries)

```{r}
# Plot the results

health_impact <- data  |> 
  group_by(EVTYPE)  |> 
  summarize(Total_Fatalities = sum(FATALITIES), Total_Injuries = sum(INJURIES))  |> 
  arrange(desc(Total_Injuries))

ggplot(health_impact[1:10, ], aes(x = reorder(EVTYPE, -Total_Injuries), y = Total_Injuries)) +
  geom_bar(stat = "identity") +
  labs(title = "Top 10 Most Harmful Events to Population Health (Injuries)",
       x = "Event Type", y = "Total Injuries") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Events with Greatest Economic Consequences

```{r}

economic_impact <- data %>%
  group_by(EVTYPE) %>%
  summarize(Total_Economic_Damage = sum(totalEconomicDmg)) %>%
  arrange(desc(Total_Economic_Damage))

# Plot the results
ggplot(economic_impact[1:10, ], aes(x = reorder(EVTYPE, -Total_Economic_Damage), y = Total_Economic_Damage)) +
  geom_bar(stat = "identity") +
  labs(title = "Top 10 Events with Greatest Economic Consequences",
       x = "Event Type", y = "Total Economic Damage (in dollars)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Conclusions

As we see, Tornado is the most harmful weather event. It is important that authorities help to solve this situation.



